---
pagetitle: 'Exercise: Environmental Stochasticity'
output: 
  html_document:
    toc: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy(position = c('top', 'right'))
```

<!--
Implement to improvements to the below code.
(i) Draw the random values for lambda inside the for loop, that is one new value at each iteration of the loop. I think this is more intuitive.
(ii) Use the "<<-" command to export time series produced by a function directly to the global environment. This should slightly simplify the code.
-->

# Environmental Stochasticity {.tabset}

## Environmental stochasticity

In the lecture, we derived

$$
N(t)=N(0)*(1+\beta-\delta)^t=N(0)*\lambda^t
$$
as a model for exponential growth in discrete time, where $\lambda = 1 + \beta - \delta$.
<!--If the population size is sufficiently large, this becomes a deterministic model because the number of offspring produced, *B*, and the number of individuals dying, *D*, at any time step will be very close to the expected values $N(t)\beta$ and $N(t)\delta$, respectively. This is due to the [law of large numbers](https://en.wikipedia.org/wiki/Law_of_large_numbers). We will investigate the effect of *demographic stochasticity* as it occurs in small populations in a later exercise.-->
Importantly, the above equation assumes that the growth factor $\lambda$ does not vary in time. In this exercise, we will relax this assumption. For example, climatic conditions vary among years and this might affect the average number of offspring, $\beta$, and the probability to die, $\delta$, of ALL individuals in a population. In short, $\lambda$ will vary in time and we indicate this by adding $t$ as an argument, $\lambda(t)$.

In this exercise, we simulate environmental stochasticity in *R*. We do this for the simplest possible case, namely, that $\lambda(t)$ can take one of two different values:
$$
\lambda_1 = \bar \lambda + \epsilon\\
\lambda_2 = \bar \lambda - \epsilon
$$
For simplicity, let us furthermore assume that each environment occurs with equal probability. We can do this in *R* by using the function ```sample(x = possible_outcomes, size = tmax, prob = probabilities, replace = TRUE)```:

```{r eval=FALSE}
lambda_mean <- 1.01
deviation <- 0.2
lambda1 <- (lambda_mean + deviation)
lambda2 <- (lambda_mean - deviation)
possible_outcomes <- c(lambda1, lambda2)
probabilities <- c(0.5, 0.5)
N0 <- 10 # initial population size
tmax <- 200 # length of the time series

sample(x = possible_outcomes, size = tmax, prob = probabilities, replace = TRUE) # produces a vector of tmax random numbers taking the value 0.21  1.21
```

In the next step, we want to use the sequence of $\lambda$-values to project a population that consists of *N0* individuals at time  *t=0* *t* time steps into the future. Thus, we want to compute

$$
N(t)=\lambda(t-1)*\lambda(t-2)*\ldots*\lambda(2)*\lambda(1)*\lambda(0)*N(0).
$$

In the following, we define a function that produces a stochastic time series.

```{r eval=FALSE}
environmental_stochasticity <- function(N0, tmax, possible_outcomes, probabilities) {
  time_series <- c()
  time_series[1] <- N0
  for (t in 1:tmax) {
    lambda <- sample(possible_outcomes, 1, probabilities, replace = TRUE)
    time_series[t+1] <- time_series[t]*lambda
  }
  return(time_series)
}
```

Next, we plot the resulting time series.

```{r eval=FALSE}
time_series <- environmental_stochasticity(N0, tmax, possible_outcomes, probabilities) # runs the function

plot(time_series, type="l", xlab="time", ylab="population size")
```

Alternatively, we can also plot the time series on a log-scale by taking the natural logarithm of population size.

```{r eval=FALSE}
plot(log(time_series + 1), type="l", xlab="time", ylab="ln(population size)")
```

Note that for the log-scale we add 1 to all entries in the time series. The reason is that $\ln(N)$ for $N<1$ is negative.

Next, let us add a line that shows how the population changes if each year it grows according to the arithmetic mean growth factor $\bar\lambda$. The following piece of code adds the desired line, both on the natural and log-scale

```{r eval=FALSE}
time_series_geom <- c()
time_series_geom[1] <- N0
for (t in 1:tmax) {
  time_series_geom[t+1] <- time_series_geom[t]*lambda_mean
}

# plots on natural scale
plot(time_series, type="l", ylim=c(0, max(time_series, time_series_geom)), xlab="time", ylab="population size")
lines(time_series_geom, type="l", col="Red")

# plots on log-scale
plot(log(time_series +1), type="l", ylim=c(0, log(max(time_series, time_series_geom))), xlab="time", ylab="ln(population size)")
lines(log(time_series_geom + 1), type="l", col="Red")
```

At each instance this code is executed, we obtain a different random time series. If you produce several time series, you will note that in some cases the population decreases to very low values. In real life, this might mean that a population goes extinct. To incorporate this into our simulations, we can introduce an *extinction threshold*. If the population size drops below this threshold, then the population is extinct and cannot recover. A natural choice for such a threshold is 1. In the following code, we thus set population size to zero whenever *N(t)* drops below 1:

```{r eval=FALSE}
environmental_stochasticity_ext <- function(N0, tmax, possible_outcomes, probabilities) {
  time_series <- c()
  time_series[1] <- N0
  for (t in 1:tmax) {
    lambda <- sample(possible_outcomes, 1, probabilities, replace = TRUE)
    time_series[t+1] <- ifelse(time_series[t] * lambda < 1, 0, time_series[t] * lambda) # sets the population vector to 0 in case N drops below 1
  }
  return(time_series)
}

time_series_ext <- environmental_stochasticity_ext(N0, tmax, possible_outcomes, probabilities) # runs the function
N0
plot(time_series_ext, type="l", xlab="time", ylab="population size")
lines(time_series_geom, type="l", col="Red")

plot(log(time_series_ext + 1), type="l", xlab="time", ylab="ln(population size)") # same data on a log-scale
lines(log(time_series_geom+1), type="l", col="Red")
```
You might encounter simulation runs in which the population goes extinct while the projection based on the geometric mean suggests that the population should slowly increase. How can we explain this apparent contradiction?

## The probability of extinction

You have seen that for the chosen set of parameters $\bar \lambda$ and $\epsilon$ a certain proportion of simulation runs results in extinction. How does the probability of extinction depend on these parameters? Let us explore this question by changing the value of $\epsilon$.

To investigate this question more systematically, it is useful to automatize the process in *R*, so that we can produce many time series in one step. For this part, we use the function with the extinction threshold and plot the time series on a log scale.

Specifically, the next piece of code allows us to execute the population simulation several times by putting our function inside a *for*-loop. How often the times series is repeated is determined by the new parameter *sim*, below set equal to 10. Note that each time series is different from the others because it is based on a different set of random $\lambda$-values. 

```{r eval=FALSE}
sim <- 10 # number of simulation runs

results <- matrix(nrow = tmax + 1, ncol = sim) # creates an empty matrix with n+1 rows and sim columns in which we will store the time series

for (i in 1:sim) {# executes the function sim times
  results[, i] <- environmental_stochasticity(N0, tmax, possible_outcomes, probabilities) # produces a time series and stores it in the jth column of the matrix results
}
```

Let us combine all time series into a single plot.

```{r eval=FALSE}
cl <- rainbow(sim) # set number of colors equal to number of simulations

plot(log(results[, 1] + 1), type="l", ylim=c(0, log(max(time_series_geom, results))), col=cl[1], xlab="time", ylab="ln(population size)") # plot the result from the first simulation
for (i in 2:sim) {# add the results from all other simulations to the same graph
  lines(log(results[, i] + 1), type="l", col=cl[i])
}
lines(log(time_series_geom + 1), type="l", col="4")
```


```{r eval=FALSE}
log_row_means <- rowMeans(log(results + 1))

# we can add mean of all time series to the previous graph
lines(log_row_means, type="l", col="Black")
```

```{r eval=FALSE, echo=FALSE}
# code for making plots for the script

lambda_mean <- 1.01
deviation <- 0.2
lambda1 <- (lambda_mean + deviation)
lambda2 <- (lambda_mean - deviation)
possible_outcomes <- c(lambda1, lambda2)
probabilities <- c(0.5, 0.5)
N0 <- 10 # initial population size
tmax <- 200 # length of the time series

environmental_stochasticity <- function(N0, tmax, possible_outcomes, probabilities) {
  growth_factors <- sample(possible_outcomes, tmax, probabilities, replace = TRUE)
  time_series <- c()
  time_series[1] <- N0
  for (t in 1:tmax) {
    time_series[t+1] <- time_series[t]*growth_factors[t]
  }
  return(time_series)
}

sim <- 40 # number of simulation runs

results <- matrix(nrow = tmax + 1, ncol = sim)

for (i in 1:sim) {# executes the function sim times
  results[, i] <- environmental_stochasticity(N0, tmax, possible_outcomes, probabilities) # produces a time series and stores it in the jth column of the matrix results
}

cl <- rainbow(sim)
plot(log(results[, 1] + 1), type="l", ylim=c(0, log(max(time_series_geom, results) + 1)), col=cl[1], xlab="time", ylab="ln(population size)")
for (i in 2:sim) {
  lines(log(results[, i] + 1), type="l", col=cl[i])
}
lines(log(time_series_geom + 1), type="l", col="4")

log_row_means <- rowMeans(log(results + 1))
lines(log_row_means, type="l", col="Black")

exp_mean_lambda <- sqrt(lambda1*lambda2) # exponential of the formula in the above equation

time_series_expected <- c()
time_series_expected[1] <- N0
for (t in 1:tmax) {
  time_series_expected[t+1] <- time_series_expected[t]*exp_mean_lambda
}

# we can add the expected mean time series to the previous graph
lines(log(time_series_expected + 1), type="l", col="Black")
```

For the environmental variation investigated here (with only two possible environments), one can show that the mean over many trajectories is given by

$$
\begin{equation}
\bar N(t) = N_0 \exp(\bar r t),
\end{equation}
$$
where

$$
\begin{equation}\label{eq:bar_r}
\bar r = \ln \bar\lambda + \frac{1}{2} \ln \left(1 - \frac{\epsilon^2}{\bar\lambda^2}\right).
\end{equation}
$$
Let us add a line showing $\bar N(t)$ according to this formula to our plots.

```{r eval=FALSE}
exp_mean_lambda <- exp(log(lambda_mean) + 1/2*log(1 - deviation^2 / lambda_mean^2)) # exponential of the formula in the above equation

time_series_expected <- c()
time_series_expected[1] <- N0
for (t in 1:tmax) {
  time_series_expected[t+1] <- time_series_expected[t]*exp_mean_lambda
}

# we can add the expected mean time series to the previous graph
lines(log(time_series_expected + 1), type="l", col="Gray")
```

```{r eval=FALSE, echo=FALSE}
## code used for figures for lecture notes
# first, simulations without extinction threshold (both log- and normal scale)
####################
lambda_mean <- 1.01
deviation <- 0.2
lambda1 <- (lambda_mean + deviation)
lambda2 <- (lambda_mean - deviation)
possible_outcomes <- c(lambda1, lambda2)
probabilities <- c(0.5, 0.5)
N0 <- 10 # initial population size
tmax <- 200 # length of the time series

environmental_stochasticity <- function(N0, tmax, possible_outcomes, probabilities) {
  growth_factors <- sample(possible_outcomes, tmax, probabilities, replace = TRUE)
  time_series <- c()
  time_series[1] <- N0
  for (t in 1:tmax) {
    time_series[t+1] <- time_series[t]*growth_factors[t]
  }
  return(time_series)
}

sim <- 40 # number of simulation runs

results <- matrix(nrow = tmax + 1, ncol = sim)

for (i in 1:sim) {# executes the function sim times
  results[, i] <- environmental_stochasticity(N0, tmax, possible_outcomes, probabilities) # produces a time series and stores it in the jth column of the matrix results
}

time_series_geom <- c()
time_series_geom[1] <- N0
for (t in 1:tmax) {
  time_series_geom[t+1] <- time_series_geom[t]*lambda_mean
}

cl <- rainbow(sim)

plot(log(results[, 1] + 1), type="l", ylim=c(0, log(max(time_series_geom, results) + 1)), col=cl[1], xlab="time", ylab="ln(population size)")
for (i in 2:sim) {
  lines(log(results[, i] + 1), type="l", col=cl[i])
}

lines(log(time_series_geom + 1), type="l", col="4")

log_row_means <- rowMeans(log(results + 1))
lines(log_row_means, type="l", col="Black")

exp_mean_lambda <- sqrt(lambda1*lambda2) # exponential of the formula in the above equation

time_series_expected <- c()
time_series_expected[1] <- N0
for (t in 1:tmax) {
  time_series_expected[t+1] <- time_series_expected[t]*exp_mean_lambda
}

lines(log(time_series_expected + 1), type="l", col="Black")

# plot on normal scale
plot(results[, 1], type="l", ylim=c(0, max(time_series_geom, results)), col=cl[1], xlab="time", ylab="population size")
for (i in 2:sim) {
  lines(results[, i], type="l", col=cl[i])
}

lines(time_series_geom, type="l", col="4")

log_row_means <- rowMeans(results)
lines(log_row_means, type="l", col="Black")

exp_mean_lambda <- sqrt(lambda1*lambda2) # exponential of the formula in the above equation

time_series_expected <- c()
time_series_expected[1] <- N0
for (t in 1:tmax) {
  time_series_expected[t+1] <- time_series_expected[t]*exp_mean_lambda
}

lines(time_series_expected, type="l", col="Black")

####################
# second, simulations WITH extinction threshold (only log-scale)
####################

lambda_mean <- 1.01
deviation <- 0.05
lambda1 <- (lambda_mean + deviation)
lambda2 <- (lambda_mean - deviation)
possible_outcomes <- c(lambda1, lambda2)
probabilities <- c(0.5, 0.5)
N0 <- 10 # initial population size
tmax <- 200 # length of the time series

environmental_stochasticity_ext <- function(N0, tmax, possible_outcomes, probabilities) {
  growth_factors <- sample(possible_outcomes, tmax, probabilities, replace = TRUE)
  time_series <- c()
  time_series[1] <- N0
  for (t in 1:tmax) {
    time_series[t+1] <- ifelse(time_series[t]*growth_factors[t] < 1, 0, time_series[t]*growth_factors[t]) # sets the population vector to 0 in case N drops below 1
  }
  return(time_series)
}

sim <- 40 # number of simulation runs

results <- matrix(nrow = tmax + 1, ncol = sim)

for (i in 1:sim) {# executes the function sim times
  results[, i] <- environmental_stochasticity_ext(N0, tmax, possible_outcomes, probabilities) # produces a time series and stores it in the jth column of the matrix results
}

time_series_geom <- c()
time_series_geom[1] <- N0
for (t in 1:tmax) {
  time_series_geom[t+1] <- time_series_geom[t]*lambda_mean
}

cl <- rainbow(sim)

plot(log(results[, 1] + 1), type="l", ylim=c(0, log(max(time_series_geom, results) + 1)), col=cl[1], xlab="time", ylab="ln(population size)")
for (i in 2:sim) {
  lines(log(results[, i] + 1), type="l", col=cl[i])
}

lines(log(time_series_geom + 1), type="l", col="4")

#log_row_means <- rowMeans(log(results + 1))
#lines(log_row_means, type="l", col="Black")

exp_mean_lambda <- sqrt(lambda1*lambda2) # exponential of the formula in the above equation

time_series_expected <- c()
time_series_expected[1] <- N0
for (t in 1:tmax) {
  time_series_expected[t+1] <- time_series_expected[t]*exp_mean_lambda
}

lines(log(time_series_expected + 1), type="l", col="Black")
```

### TASKS

From our graphs, one can determine the number of simulation runs in which the population goes extinct, simply by counting the number of simulations that hit zero.

  (1)
  In your latest plot, you have combined four types of line: (i) simulated stochastic time series, (ii) the mean over these simulated time series, (iii) the mathematically calculated *expected* mean trajectory, and (iv) the trajectory predicted by the arithmetic mean growth factor $\bar\lambda$.
Summerize what you learned from these graphs. What is the effect of environmental stochasticity on population growth and population survival?
  
  (2)
  Investigate the magnitude of environmental fluctuations in more detail by varying the amplitude of the environmental fluctuations. Keep $\bar \lambda = 1.01$ fixed and vary $\epsilon$, the deviation in $\lambda$, according to
$\epsilon = (0; 0.05; 0.1; 0.15; 0.2)$.
For each value of $\epsilon$ report the number of simulations that result in extinction based on **20** simulation runs. Run you simulations for at least **300** time steps.
For which value of $\epsilon$ does the expected mean trajectory starts to decline?

  (3)
Submit your results.

