---
pagetitle: 'Exercise: Environmental Stochasticity'
output: 
  html_document:
    toc: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy(position = c('top', 'right'))
```

<!--
(i) Use the "<<-" command to export time series produced by a function directly to the global environment. This should slightly simplify the code.
-->

# Environmental Stochasticity {.tabset}

## Environmental stochasticity

In the lecture, we derived

$$
N(t)=N(0)*(1+\beta-\delta)^t=N(0)*\lambda^t
$$
as a model for exponential growth in discrete time, where $\lambda = 1 + \beta - \delta$.
<!--If the population size is sufficiently large, this becomes a deterministic model because the number of offspring produced, *B*, and the number of individuals dying, *D*, at any time step will be very close to the expected values $N(t)\beta$ and $N(t)\delta$, respectively. This is due to the [law of large numbers](https://en.wikipedia.org/wiki/Law_of_large_numbers). We will investigate the effect of *demographic stochasticity* as it occurs in small populations in a later exercise.-->
Importantly, the above equation assumes that the growth factor $\lambda$ does not vary in time. In this exercise, we will relax this assumption. For example, climatic conditions vary among years and this might affect the average number of offspring, $\beta$, and the probability to die, $\delta$, of ALL individuals in a population. In short, $\lambda$ will vary in time and we indicate this by adding $t$ as an argument, $\lambda(t)$.

In this exercise, we simulate environmental stochasticity in *R*. We do this for the simplest possible case, namely, that $\lambda(t)$ can take one of two different values:
$$
\lambda_1 = \bar \lambda + \epsilon\\
\lambda_2 = \bar \lambda - \epsilon
$$
For simplicity, let us furthermore assume that each environment occurs with equal probability. We can do this in *R* by using the function ```sample(x = possible_outcomes, size = tmax, prob = probabilities, replace = TRUE)```:

```{r eval=FALSE}
lambda.mean <- 1.01
deviation <- 0.2
lambda1 <- (lambda.mean + deviation)
lambda2 <- (lambda.mean - deviation)
possible.outcomes <- c(lambda1, lambda2)
probabilities <- c(0.5, 0.5)
N0 <- 10 # initial population size
tmax <- 200 # length of the time series

sample(x = possible.outcomes, size = tmax, prob = probabilities, replace = TRUE) # example of stochastic sequence of lambda-values
```
In the next step, we want to use the sequence of $\lambda$-values to project a population that consists at time  *t=0* of *N(0)* individuals *t* time steps into the future. Thus, we want to compute

$$
N(t)=\lambda(t-1)*\lambda(t-2)*\ldots*\lambda(2)*\lambda(1)*\lambda(0)*N(0).
$$

In the following, we define a function that produces a stochastic time series.

```{r eval=FALSE}
environmental.stochasticity <- function(N0, tmax, possible.outcomes, probabilities) {
  time.series <- c(N0) # creates a vector in which the time series will be stored and enters the starting value in the first position of that vector
  for (t in 1:tmax) {# for loop that iterates the population dynamics tmax times
    lambda <- sample(possible.outcomes, 1, probabilities, replace = TRUE) # new lambda value is chosen randomly
    time.series[t+1] <- time.series[t]*lambda
  }
  return(time.series)
}
```

Next, we plot the resulting time series.

```{r eval=FALSE}
data <- environmental.stochasticity(N0, tmax, possible.outcomes, probabilities) # runs the function

plot(data, type="l", xlab="time", ylab="population size")
```

Alternatively, we can plot the time series on a log-scale.

```{r eval=FALSE}
plot(log(data+1), type="l", xlab="time", ylab="ln(population size)")
```

Note that for the log-scale we add 1 to all entries in the time series. The reason is that $\ln(N)$ for $N<1$ is negative.

Each time this code is executed, you obtain a different random time series. If you do this often enough, you will note that in some cases the population decreases to very low values. In real life, this might mean that a population goes extinct. To incorporate this into our simulations, we can introduce an *extinction threshold*. If the population size drops below this threshold, then the population is extinct and cannot recover. A natural choice for such a threshold is 1. In the following code, we thus set population size to zero whenever *N(t)* drops below 1:

```{r eval=FALSE}
# new function that incorporates extinction
environmental.stochasticity.ext <- function(N0, tmax, possible.outcomes, probabilities) {
  time.series <- c(N0)
  for (t in 1:tmax) {
    lambda <- sample(possible.outcomes, 1, probabilities, replace = TRUE)
    time.series[t+1] <- ifelse(time.series[t] * lambda < 1, 0, time.series[t] * lambda) # sets population size to 0 in case N drops below 1
  }
  return(time.series)
}

data <- environmental.stochasticity.ext(N0, tmax, possible.outcomes, probabilities)

plot(data, type="l", xlab="time", ylab="population size")

plot(log(data+1), type="l", xlab="time", ylab="ln(population size)") # same data on a log-scale
```

Let us investigate the possibility of extinction more systematically by plotting many time series in the same graph. This can be done by putting the function inside a *for*-loop. How often the times series is repeated is determined by the new parameter *sim*. 

```{r eval=FALSE}
sim <- 40 # number of simulation runs

results <- matrix(nrow = tmax + 1, ncol = sim) # Creates an empty matrix with n+1 rows and as many columns as determined by the parameter "sim". We will store each time series in one column of this matrix. 

for (i in 1:sim) {# executes the function sim times
  results[, i] <- environmental.stochasticity.ext(N0, tmax, possible.outcomes, probabilities) # produces a time series and stores it in the i-th column of the matrix results
}

# Let us combine all time series into a single plot.

cl <- rainbow(sim) # set number of colors equal to number of simulations

# next, we use the command "matplot", which plots all columns of a matrix as a separate line
matplot(results, type="l", lty=1, ylim=c(0, max(results)), col=cl, xlab="time", ylab="population size") # plots the result from the first simulation
```

You probably obtained a figure in which you cannot distinguish the different simulations very well since many of them have similarly low values. It is therefore useful view these time series on a log-scale.

```{r eval=FALSE}
matplot(log(results+1), lty = 1, type = "l", ylim = c(0, max(log(results))), col = cl, xlab = "time", ylab = "ln(population size)") # plots the result from the first simulation on a log-scale
```

Based on your simulations, you should be able to make the following observations:

1.
Each realization is different (because simulations are *stochastic*).
2.
Some simulation result in extinction while others do not.
3.
The variance across simulation runs increases with time. Thus, predictions become less accurate the longer we look into the future.
4.
The endpoints of the different simulations are highly skewed, with a few simulations reaching very high values and many quite low values.

## Average population trajectory

You have seen that as time proceeds, the different simulation runs increasingly differ from each other and some time series go extinct. But what happens to the average over all time series? This average is given by the mean over all time series $N_i(t) at each point in time, defined as
$$
\bar N(t) = \frac{N_1(t)+ \ldots + N_n(t)}{n} = \frac{\sum_{i=1}^n N_i(t)}{n}.
$$
Let us add this average to our plot. First, we repeat the code that produces *sim=10* time series with an extinction threshold on a log-scale.

```{r eval=FALSE}
sim <- 10

results <- matrix(nrow = tmax + 1, ncol = sim)

for (i in 1:sim) {
  results[, i] <- environmental.stochasticity.ext(N0, tmax, possible.outcomes, probabilities)
}

# Let us combine all time series into a single plot.

cl <- rainbow(sim)

matplot(log(results+1), lty = 1, type = "l", ylim = c(0, log(max(results)+1)), col = cl, xlab = "time", ylab = "ln(population size)")
```

To add the average is easy. In the above code, we stored the time series in a matrix called *results*. Each column contains a different time series. To calculate the mean population size at a certain point in time we have to take the mean over all values within a row. This can be done with the command ```rowMeans```.

```{r eval=FALSE}
row.means <- rowMeans(log(results + 1)) # important to first add 1, then take the logarithm and average in last step

lines(row.means, type="l", lwd = 2, col="Black")
```

You should now see a black wiggly line appearing in you plot. This line should become smoother the higher the number of simulations (try it).

Is there a way to predict the average trajectory? It might be tempting to think that the average should follow
$$
\bar N(t) = N(0) \bar \lambda^t,
$$
where $\bar \lambda = 1/t*\sum_{i=0}^{t-1} \lambda(i)$ is the \textit{arithmetic mean} of the annual growth factors. In our current scenario, we have $\bar\lambda =$*lambda.mean*. Let us add this line.

```{r eval=FALSE}
# function computing a time series given that the population changes by the arithmetic mean lambda value
time.series.arith <- c(N0)
for (t in 1:tmax) {
  time.series.arith[t+1] <- time.series.arith[t]*lambda.mean
}

# adds a line with the time series "time.series.arith" to the figure
lines(log(time.series.arith+1), type="l", lwd = 2, col="4")
```

You should now see a straight blue line showing exponential growth on a log-scale. It shows how the population should grow if it would change by the factor $\bar \lambda$ from one year to the next. Well, this line does not seem to do a good job in predicting the average population trajectory (the black wiggly line).

Next, let us add a line that shows how the time series develops if the population would grow according the the *geometric mean* $\bar\lambda_{\mathrm G}$ from one year to the next. Given our current assumptions of two possible environmental states (*good* and *bad* years), each occurring with probability $1/2$, we have
$$
\bar \lambda_{\mathrm G} = \sqrt{\lambda_1 \lambda_2} = \left((\bar \lambda + \epsilon)(\bar \lambda - \epsilon) \right)^\frac{1}{2} = \left(\bar \lambda^2  - \epsilon^2 \right)^\frac{1}{2}
$$

```{r eval=FALSE}
# function computing a time series given that the population changes by the geometric mean lambda value
lambda.geo <- (lambda.mean^2-deviation^2)^(1/2) # calculates the geometric mean lambda value according to the above formula
lambda.geo
time.series.geo <- c(N0)
for (t in 1:tmax) {
  time.series.geo[t+1] <- time.series.geo[t]*lambda.geo
}

# adds a line with the time series "time.series.geo" to the figure
lines(log(time.series.geo+1), type="l", lwd = 2, col="Black")
```

This new smooth black line much better approximates the average trajectory (given by the wiggly black line). Indeed, one can prove that the trajectory based on a very large number of simulations is given by
$$
\begin{equation}
\bar N(t) = N(0) \bar \lambda_{\mathrm G}^t.
\end{equation}
$$
Note that this statement is true for the version of the model without extinction threshold. Furthermore, note that in the case with many decreasing trajectories, we see a larger difference between the average trajectory and trajectory based on the geometric mean. This is because if trajectories reach very low numbers, then the fact that we add +1 to our data becomes visible. 

## Probability of extinction

You have seen that for the chosen set of parameters $\bar \lambda$ and $\epsilon$ a certain proportion of simulation runs results in extinction. How does the probability of extinction depend on these parameters? Let us explore this question by changing the value of $\epsilon$. For this part, we use the function with the extinction threshold and plot the time series on a log-scale.

### TASKS

From our graphs, one can determine the number of simulation runs in which the population goes extinct, simply by counting the number of simulations that hit zero.

  (1) Investigate the magnitude of environmental fluctuations in more detail by varying the amplitude of the environmental fluctuations. Keep $\bar \lambda = 1.01$ fixed and vary $\epsilon$, the deviation from the mean $\lambda$-value, according to
$\epsilon = (0; 0.05; 0.1; 0.15; 0.2)$.
For each value of $\epsilon$, report the number of simulations that result in extinction based on **20** simulation runs. Run you simulations for at least **300** time steps.
For which value of $\epsilon$ does the expected mean trajectory starts to decline?

  (2)
Submit your work, either by using *compile report*, or, even better, as RMarkdown document compiled to HTML.

```{r eval=FALSE, echo=FALSE}
## code used for figures for lecture notes
# first, simulations without extinction threshold (both log- and normal scale)
####################
lambda.mean <- 1.01
deviation <- 0.2
lambda1 <- (lambda.mean + deviation)
lambda2 <- (lambda.mean - deviation)
possible.outcomes <- c(lambda1, lambda2)
probabilities <- c(0.5, 0.5)
N0 <- 10 # initial population size
tmax <- 200 # length of the time series

environmental.stochasticity <- function(N0, tmax, possible.outcomes, probabilities) {
  time.series <- c()
  time.series[1] <- N0
  for (t in 1:tmax) {
    lambda <- sample(possible.outcomes, 1, probabilities, replace = TRUE)
    time.series[t+1] <- time.series[t] * lambda
  }
  return(time.series)
}

sim <- 40 # number of simulation runs

results <- matrix(nrow = tmax + 1, ncol = sim)

for (i in 1:sim) {# executes the function sim times
  results[, i] <- environmental.stochasticity(N0, tmax, possible.outcomes, probabilities) # produces a time series and stores it in the jth column of the matrix results
}

time.series.geom <- c()
time.series.geom[1] <- N0
for (t in 1:tmax) {
  time.series.geom[t+1] <- time.series.geom[t]*lambda.mean
}

cl <- rainbow(sim)

plot(log(results[, 1] + 1), type="l", ylim=c(0, log(max(time.series.geom, results) + 1)), col=cl[1], xlab="time", ylab="ln(population size)")
for (i in 2:sim) {
  lines(log(results[, i] + 1), type="l", col=cl[i])
}

lines(log(time.series.geom + 1), type="l", col="4")

log.row.means <- rowMeans(log(results + 1))
lines(log.row.means, type="l", col="Black")

exp.mean.lambda <- sqrt(lambda1*lambda2) # exponential of the formula in the above equation

time.series.expected <- c()
time.series.expected[1] <- N0
for (t in 1:tmax) {
  time.series.expected[t+1] <- time.series.expected[t]*exp.mean.lambda
}

lines(log(time.series.expected + 1), type="l", col="Black")

# plot on normal scale
plot(results[, 1], type="l", ylim=c(0, max(time.series.geom, results)), col=cl[1], xlab="time", ylab="population size")
for (i in 2:sim) {
  lines(results[, i], type="l", col=cl[i])
}

lines(time.series.geom, type="l", col="4")

log.row.means <- rowMeans(results)
lines(log.row.means, type="l", col="Black")

exp.mean.lambda <- sqrt(lambda1*lambda2) # exponential of the formula in the above equation

time.series.expected <- c()
time.series.expected[1] <- N0
for (t in 1:tmax) {
  time.series.expected[t+1] <- time.series.expected[t]*exp.mean.lambda
}

lines(time.series.expected, type="l", col="Black")

####################
# second, simulations WITH extinction threshold (only log-scale)
####################

lambda.mean <- 1.01
deviation <- 0.05
lambda1 <- (lambda.mean + deviation)
lambda2 <- (lambda.mean - deviation)
possible.outcomes <- c(lambda1, lambda2)
probabilities <- c(0.5, 0.5)
N0 <- 10 # initial population size
tmax <- 200 # length of the time series

environmental.stochasticity.ext <- function(N0, tmax, possible.outcomes, probabilities) {
  growth.factors <- sample(possible.outcomes, tmax, probabilities, replace = TRUE)
  time.series <- c()
  time.series[1] <- N0
  for (t in 1:tmax) {
    time.series[t+1] <- ifelse(time.series[t]*growth.factors[t] < 1, 0, time.series[t]*growth.factors[t]) # sets the population vector to 0 in case N drops below 1
  }
  return(time.series)
}

sim <- 40 # number of simulation runs

results <- matrix(nrow = tmax + 1, ncol = sim)

for (i in 1:sim) {# executes the function sim times
  results[, i] <- environmental.stochasticity.ext(N0, tmax, possible.outcomes, probabilities) # produces a time series and stores it in the jth column of the matrix results
}

time.series.geom <- c()
time.series.geom[1] <- N0
for (t in 1:tmax) {
  time.series.geom[t+1] <- time.series.geom[t]*lambda.mean
}

cl <- rainbow(sim)

plot(log(results[, 1] + 1), type="l", ylim=c(0, log(max(time.series.geom, results) + 1)), col=cl[1], xlab="time", ylab="ln(population size)")
for (i in 2:sim) {
  lines(log(results[, i] + 1), type="l", col=cl[i])
}

lines(log(time.series.geom + 1), type="l", col="4")

#log.row.means <- rowMeans(log(results + 1))
#lines(log.row.means, type="l", col="Black")

exp.mean.lambda <- sqrt(lambda1*lambda2) # exponential of the formula in the above equation

time.series.expected <- c()
time.series.expected[1] <- N0
for (t in 1:tmax) {
  time.series.expected[t+1] <- time.series.expected[t]*exp.mean.lambda
}

lines(log(time.series.expected + 1), type="l", col="Black")
```

```{r eval=FALSE, echo=FALSE}
# code for making plots for the lecture notes

lambda.mean <- 1.01
deviation <- 0.2
lambda1 <- (lambda.mean + deviation)
lambda2 <- (lambda.mean - deviation)
possible.outcomes <- c(lambda1, lambda2)
probabilities <- c(0.5, 0.5)
N0 <- 10 # initial population size
tmax <- 200 # length of the time series

environmental.stochasticity <- function(N0, tmax, possible.outcomes, probabilities) {
  growth.factors <- sample(possible.outcomes, tmax, probabilities, replace = TRUE)
  time.series <- c()
  time.series[1] <- N0
  for (t in 1:tmax) {
    time.series[t+1] <- time.series[t]*growth.factors[t]
  }
  return(time.series)
}

sim <- 40 # number of simulation runs

results <- matrix(nrow = tmax + 1, ncol = sim)

for (i in 1:sim) {# executes the function sim times
  results[, i] <- environmental.stochasticity(N0, tmax, possible.outcomes, probabilities) # produces a time series and stores it in the jth column of the matrix results
}

cl <- rainbow(sim)
plot(log(results[, 1] + 1), type="l", ylim=c(0, log(max(time.series.geom, results) + 1)), col=cl[1], xlab="time", ylab="ln(population size)")
for (i in 2:sim) {
  lines(log(results[, i] + 1), type="l", col=cl[i])
}
lines(log(time.series.geom + 1), type="l", col="4")

log.row.means <- rowMeans(log(results + 1))
lines(log.row.means, type="l", col="Black")

exp.mean.lambda <- sqrt(lambda1*lambda2) # exponential of the formula in the above equation

time.series.expected <- c()
time.series.expected[1] <- N0
for (t in 1:tmax) {
  time.series.expected[t+1] <- time.series.expected[t]*exp.mean.lambda
}

# we can add the expected mean time series to the previous graph
lines(log(time.series.expected + 1), type="l", col="Black")
```